Supplementary materials: Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-10-10 docs/manuscript.Rmd
updated

1. $H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0$; *the SNP associates with the exposure*
2. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
4. $H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*

where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. The 4th condition is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate against from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]). We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders of the exposure and outcome are present, as these will induce collider bias.

@@

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then model 1 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then model 2 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is likely confounding and no call is made ($x \leftarrow u \rightarrow y$).

@@

Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Supplementary text 3). The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (Supplementary figure 1). Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.

@@

**Supplementary figure 1** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

>>>>

1. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
2. $H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0$; *the SNP associates with the exposure conditional on the outcome*
3. $H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0$; *the exposure associates with the outcome conditional on the SNP*
4. $H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0$; *the SNP is independent of the outcome conditional on the exposure*

The term in the 4th test can be rewritten as a $cov(g, y|x) = cov(g, y - \hat{y})$ where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, and $x$ is assumed to mediate the association between the SNP and the outcome. The condition in the 4th test is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]). We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders of the exposure and outcome are present, as these will induce collider bias.

@@

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then model 1 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then model 2 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is potentially confounding (Supplementary figure 1) and no call is made.

@@

Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Supplementary text 3). The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (Supplementary figure 2). Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.

@@

**Supplementary figure 1** Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding. Here, the phenotypes $x$ and $y$ are not causally related, but there is a genetic effect and a confounder both influencing each phenotype. Each point represents a single simulation. Where power is high (when the absolute values of the $x$ and $y$ axes are large) the CIT returns a significant result ($p < 0.01$) when testing the causal effect of $x$ on $y$, and when testing the causal effect of $y$ on $x$.

**Supplementary figure 2** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-10 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{28 July 2017}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0\); \emph{the SNP
  associates with the exposure}
\item
  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
  associates with the outcome}
\item
  \(H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0\); \emph{the exposure
  associates with the outcome}
\item
  \(H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0\);
  \emph{the SNP is independent of the outcome when the outcome is
  adjusted for the exposure}
\end{enumerate}

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. The 4th
condition is formulated as an equivalence testing problem that is
estimated using simulations, comparing the estimate against from the
data against empirically obtained estimates for simulated variables
where the independence model is true (full details are given in (4)). We
note here that this approach is liable to fail, even when there is a
true causal relationship, when confounders of the exposure and outcome
are present, as these will induce collider bias.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then there is likely confounding
  and no call is made (\(x \leftarrow u \rightarrow y\)).
\end{itemize}

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 2 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=\rho_{y, y_o}1\))
the CIT is reliable in identifying the correct causal direction.
However, as measurement error increases in the exposure variable,
eventually the CIT is more likely to infer a robust causal association
in the wrong direction. Also of concern here is that increasing sample
size does not solve the issue, indeed it only strengthens the apparent
evidence for the incorrect inference.

@@

Figure 3: Outcomes were simulated to be unrelated to the exposure
(bottom plot) or causally influenced by the exposure (top plot) with
varying degrees of measurement imprecision applied to the exposure
variable (x axis). True positive rates (y axis) for MR and CIT were
compared for varying sample sizes (columns of boxes).

@@

Figure 4: a) We can predict the values the MR Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by the range of values where the blue surface is above the
red plane, those values of measurement error lead to our observed MR
Steiger test inferring the wrong causal direction. Where the blue
surface lies below the red plane, these measurement error values support
the inferred causal direction of X to Y. A measure of reliability,
therefore, is the ratio of the negative and positive volumes of the
total space bound by the blue and red surfaces,
\(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\). In this case, where
\(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the \(R = 4.40\),
which means that 4.40 times as much of the possible measurement error
values are in support of the \(x \rightarrow y\) direction of causality
than \(y \rightarrow x\). b) Plots depicting the parameter space in
which the function \(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is
negative. When \(d\) is negative the MR Steiger test is liable to infer
the wrong direction of causality. Shaded regions show the parameter
space where \(d\) is negative. The graph shows that for the majority of
the parameter space of the function, \(d\) is positive, especially where
causal relationships are relatively weak.

@@

\textbf{Supplementary figure 1} Graph representing the unmeasured
confounding parameters that will lead to the MR Steiger test returning
the wrong causal direction. Columns of boxes represent different signed
values of the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{10 October 2017}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
  associates with the outcome}
\item
  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
  associates with the exposure conditional on the outcome}
\item
  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
  associates with the outcome conditional on the SNP}
\item
  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
  independent of the outcome conditional on the exposure}
\end{enumerate}

The term in the 4th test can be rewritten as a
\(cov(g, y|x) = cov(g, y - \hat{y})\) where
\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
association between the SNP and the outcome. The condition in the 4th
test is formulated as an equivalence testing problem that is estimated
using simulations, comparing the estimate from the data against
empirically obtained estimates for simulated variables where the
independence model is true (full details are given in (4)). We note here
that this approach is liable to fail, even when there is a true causal
relationship, when confounders of the exposure and outcome are present,
as these will induce collider bias.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
  confounding (Supplementary figure 1) and no call is made.
\end{itemize}

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 2 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=\rho_{y, y_o}=1\))
the CIT is reliable in identifying the correct causal direction.
However, as measurement error increases in the exposure variable,
eventually the CIT is more likely to infer a robust causal association
in the wrong direction. Also of concern here is that increasing sample
size does not solve the issue, indeed it only strengthens the apparent
evidence for the incorrect inference.

@@

Figure 3: Outcomes were simulated to be unrelated to the exposure
(bottom plot, showing false positive rates on the y-axis) or causally
influenced by the exposure (top plot, showing true positive rates on the
y-axis) with varying degrees of measurement imprecision applied to the
exposure variable (x axis). Results for MR and CIT were compared for
varying sample sizes (columns of boxes).

@@

Figure 4: a) We can predict the values the MR Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by the range of values where the blue surface is above the
black plane, those values of measurement error lead to our observed MR
Steiger test inferring the wrong causal direction. Where the blue
surface lies below the black plane, these measurement error values
support the inferred causal direction of X to Y. A measure of
reliability, therefore, is the ratio of the negative and positive
volumes of the total space bound by the blue and black surfaces,
\(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\). In this case, where
\(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the \(R = 4.40\),
which means that 4.40 times as much of the possible measurement error
values are in support of the \(x \rightarrow y\) direction of causality
than \(y \rightarrow x\). b) Plots depicting the parameter space in
which the function \(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is
negative. When \(d\) is negative the MR Steiger test is liable to infer
the wrong direction of causality. Shaded regions show the parameter
space where \(d\) is negative. The graph shows that for the majority of
the parameter space of the function, \(d\) is positive, especially where
causal relationships are relatively weak.

@@

\textbf{Supplementary figure 1} Illustrative simulations (\(n = 5000\))
showing the results from CIT analysis under a model of confounding.
Here, the phenotypes \(x\) and \(y\) are not causally related, but there
is a genetic effect and a confounder both influencing each phenotype.
Each point represents a single simulation. Where power is high (when the
absolute values of the \(x\) and \(y\) axes are large) the CIT returns a
significant result (\(p < 0.01\)) when testing the causal effect of
\(x\) on \(y\), and when testing the causal effect of \(y\) on \(x\).

\textbf{Supplementary figure 2} Graph representing the unmeasured
confounding parameters that will lead to the MR Steiger test returning
the wrong causal direction. Columns of boxes represent different signed
values of the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).

### 2017-10-10 docs/supplementary.Rmd
updated

If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values. If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality. $V_{z}$, the total volume of the function (Supplementary figure 4), can be obtained analytically by solving:

@@

Supplementary figure 1 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects. The pattern of results were similar for different values of $\beta_{gx}$. We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.

@@

Supplementary figure 1: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

>>>>

If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values. If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality. $V_{z}$, the total volume of the function (Figure 4), can be obtained analytically by solving:

@@

Supplementary figure 2 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects. The pattern of results were similar for different values of $\beta_{gx}$. We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.

![](../images/cit_confounder2.pdf)

Supplementary figure 1: Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding. Here, the phenotypes $x$ and $y$ are not causally related, but there is a genetic effect and a confounder both influencing each phenotype. Each point represents a single simulation. Where power is high (when the absolute values of the $x$ and $y$ axes are large) the CIT returns a significant result ($p < 0.01$) when testing the causal effect of $x$ on $y$, and when testing the causal effect of $y$ on $x$.

@@

Supplementary figure 2: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-10 docs/supplementary.tex
updated

\setlength{\droptitle}{-2em}
  \title{Supplementary materials: Orienting the causal relationship between
imprecisely measured traits using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{28 July 2017}

@@

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the parameter space favours the inferred direction of
causality. \(V_{z}\), the total volume of the function (Supplementary
figure 4), can be obtained analytically by solving:

@@

Supplementary figure 1 shows the relationship between the magnitude of
the correlations between \(x\), \(y\) and the confounder \(u\) for a
range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
confounder effects. The pattern of results were similar for different
values of \(\beta_{gx}\). We note that in most cases for the parameter
values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
less than 0.2, unmeasured confounding will not incur the wrong causal
direction in the MR Steiger test.

@@

Supplementary figure 1: Graph representing the unmeasured confounding
parameters that will lead to the MR Steiger test returning the wrong
causal direction. Columns of boxes represent different signed values of
the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).

>>>>

\setlength{\droptitle}{-2em}
  \title{Supplementary materials: Orienting the causal relationship between
imprecisely measured traits using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{10 October 2017}

@@

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the parameter space favours the inferred direction of
causality. \(V_{z}\), the total volume of the function (Figure 4), can
be obtained analytically by solving:

@@

Supplementary figure 2 shows the relationship between the magnitude of
the correlations between \(x\), \(y\) and the confounder \(u\) for a
range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
confounder effects. The pattern of results were similar for different
values of \(\beta_{gx}\). We note that in most cases for the parameter
values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
less than 0.2, unmeasured confounding will not incur the wrong causal
direction in the MR Steiger test.

\includegraphics{../images/cit_confounder.pdf}

Supplementary figure 1: Illustrative simulations (\(n = 5000\)) showing
the results from CIT analysis under a model of confounding. Here, the
phenotypes \(x\) and \(y\) are not causally related, but there is a
genetic effect and a confounder both influencing each phenotype. Each
point represents a single simulation. Where power is high (when the
absolute values of the \(x\) and \(y\) axes are large) the CIT returns a
significant result (\(p < 0.01\)) when testing the causal effect of
\(x\) on \(y\), and when testing the causal effect of \(y\) on \(x\).

@@

Supplementary figure 2: Graph representing the unmeasured confounding
parameters that will lead to the MR Steiger test returning the wrong
causal direction. Columns of boxes represent different signed values of
the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).
